---
title: "Covid19JHU"
author: "Dhanushga Lionel"
date: "6/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## COVID-19 R Markdown

I will start by reading in the data from the four main .csv files from the JHU COVID-19 Github page. We will also need the library(tidyverse) package.

```{r get_jhu_data}

```
Let's read in the data and see what we have.

```{r import_data, message = FALSE}
library(tidyverse)
global_cases <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")
global_deaths <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")
US_cases <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")
US_deaths <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")
```
After looking at the global_cases, and global_deaths, I would like to tidy those datasets, and put each variable (date, cases, deaths) in their own column.

Also I don't need Lat and Long for the analysis I am planning, so I will get rid of those and rename Region and State to be more R-friendly
```{R tidy_global_data}
global_cases <- global_cases %>%
     pivot_longer(cols = -c(`Province.State`,
                           `Country.Region`, Lat, Long),
                 names_to = "date",
                 values_to = "cases") %>%
     select(-c(Lat,Long))


global_deaths <- global_deaths %>%
     pivot_longer(cols = -c(`Province.State`,
                           `Country.Region`, Lat, Long),
                 names_to = "date",
                 values_to = "deaths") %>%
     select(-c(Lat,Long))
```

Combining global_cases and global_deaths together in another variable called global. We will first need the package library(lubridate)
```{R combine_global}
library(lubridate)
global <- global_cases %>%
    full_join(global_deaths) %>%
    rename(Country_Region = `Country.Region`,
           Province_State = `Province.State`) %>%
    mutate(date=mdy(str_sub(date, 2)))
```
Look at the summary of the global data.
```{R}
summary(global)
```
Filter out rows where there are no cases. There should only be rows where cases are positive. Then check the summary.
```{R no_cases}
global <- global %>% filter(cases > 0)
summary(global)
```
Check to see if Maximum cases is possible (>33,000,000)
```{R case_greater_max}
global %>% filter(cases > 33000000)
```
Want to do the same thing for the United States. First look what is in US_cases.
```{R US_cases}
US_cases
```
Now we want to tidy up the data. We are going to pivot the columns to make it more R friendly.
```{R pivot}
US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases")
```
Need to change a couple things. 

1. Get rid of Lat and Long_
2. Change date from character to date object
3. Select Admin2 to cases

```{R clean_US_cases}
US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date=mdy(str_sub(date, 2))) %>%
  select(-c(Lat, Long_))
```
Do the same thing and clean up US_deaths
```{R clean_US_death}
US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date=mdy(str_sub(date, 2))) %>%
  select(-c(Lat, Long_))
```
Join US_cases and US_deaths into 1 data set called US
```{R Join_US}
US <- US_cases %>%
  full_join(US_deaths)
```
We have population data for the US dataset, but not for the Global dataset. To do comparative analysis, we need to get population data for the Global dataset. 

First let's Combine the Combined_Key with Province_State and Country_Region
```{R Combined_Key_Global}
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
```
Now we will add the population to the Global data set. The same github repo for the JHU files has a .csv file with the population data.
```{R UID_lookup}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
```
Need to clean it up so that Province_State is "" and not NA. This will help to get Population in the Global df. This was a workaround I found, otherwise I was getting errors.
```{r}
uid$Province_State[is.na(uid$Province_State)] <- ""
```
Join population to Global dataset
```{R join_population_global}
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, Population, Combined_Key)

global
```
Now we need to visualize the dataset. We will start with the US data set. Analyzing the US by state by state, region and date. Then we will add another column called deaths_per_mill.
```{R Visualize_US_tamil}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
US_by_state
```
Now look at total of US.
```{R US_Total_visualize}
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
US_totals
```
Filter where there were cases. X variable will be date, and the y variable will have both cases and deaths. To get a sense of the graph, it will be scaled on a log10 factor.
```{R plot_US_totals}
US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 in US", y = NULL)
```
Now we will be doing it by state. 
```{R US_State_NY}
state <- "New York" 
US_by_state %>%
  filter(Province_State == state) %>%
  filter (cases > 0) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID-19 in ", state), y = NULL)
```
Maximum us total by date and deaths
```{R}
max(US_totals$date)
max(US_totals$deaths)
```
We will now transform our data again, by adding new variables. We will be adding 2 new columns, new_cases and new_deaths.
```{R}
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

tail(US_totals %>% select(new_cases, new_deaths, everything()))
```
Now we will visualize the US new cases and new deaths. We can see it is not just flat, but is going down. 
```{R Visualize_US}
US_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) + 
  geom_point(aes(color="new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 in US", y=NULL)
```
Do the same thing for a particular state. See new cases and new deaths for New York.
```{R visualize_NY}
state <- "New York" 
US_by_state %>%
  filter(Province_State == state) %>%
  filter (cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID-19 in ", state), y = NULL)
```
Now we want to see the states with the most cases and deaths. We will group by state, and choose max deaths and cases.
```{R Worst_States}
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases), 
            population = max(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

US_state_totals %>%
  slice_min(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

US_state_totals %>%
  slice_max(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```
Now we will do some modelling. Linear modelling is predicting the values of a particular variable from other variables. We will be looking at the deaths per thousand as a function of the cases per thousand. We will then make a grid from 1 to 151 and put that into a tibble.
```{R Linear_model}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)
US_state_totals %>% slice_min(cases_per_thou)
US_state_totals %>% slice_max(cases_per_thou)
x_grid <- seq(1, 151)
new_df <- tibble(cases_per_thous = x_grid)
US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
```
Now we will plot the actual and the prediction
```{R}
US_tot_w_pred %>% 
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```